---
title: "p8105_hw2_YOURUNI"
author: "Your Name"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)
library(readxl)
```

# Problem 1

```{r p1_import, message = FALSE}
# Route8-Route11 are read as numbers by default; force them to character so pivoting works later
transit_df =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")
  ) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, vending, entrance_type, ada
  ) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

Paragraph: This dataset contains ... (describe variables). Cleaning steps: cleaned names, selected the variables listed above, converted `entry` to logical. The resulting dataset has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns. These data are **not** tidy, because route information is spread across `route1`–`route11` columns.

```{r p1_questions}
# Number of distinct stations
n_stations = transit_df |> distinct(line, station_name) |> nrow()

# Number of ADA-compliant stations
n_ada = transit_df |> filter(ada == TRUE) |> distinct(line, station_name) |> nrow()

# Proportion of entrances/exits without vending that allow entry
prop_no_vend = transit_df |> filter(vending == "NO") |> pull(entry) |> mean()
```

There are `r n_stations` distinct stations, of which `r n_ada` are ADA compliant. Among entrances/exits without vending, `r round(prop_no_vend, 3)` allow entrance.

```{r p1_routes}
transit_long =
  transit_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)

a_stations = transit_long |> filter(route_name == "A") |> distinct(line, station_name, ada)

n_a = nrow(a_stations)
n_a_ada = a_stations |> filter(ada == TRUE) |> nrow()
```

`r n_a` distinct stations serve the A train; of these, `r n_a_ada` are ADA compliant.

# Problem 2

Check the exact file name, sheet names, and column ranges in your Excel file first (`excel_sheets("data/...")`), and adjust `range` so note columns on the right are excluded.

```{r p2_import}
trash_path = "data/202409 Trash Wheel Collection Data.xlsx"

mr_df =
  read_excel(trash_path, sheet = "Mr. Trash Wheel", range = cell_cols("A:N"), na = c("", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel"
  )

prof_df =
  read_excel(trash_path, sheet = "Professor Trash Wheel", range = cell_cols("A:M"), na = c("", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(year = as.numeric(year), trash_wheel = "Professor Trash Wheel")

gwynnda_df =
  read_excel(trash_path, sheet = "Gwynnda Trash Wheel", range = cell_cols("A:L"), na = c("", "NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(year = as.numeric(year), trash_wheel = "Gwynnda")

trash_df =
  bind_rows(mr_df, prof_df, gwynnda_df) |>
  relocate(trash_wheel)
```

```{r p2_questions}
prof_weight =
  trash_df |> filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |> sum(na.rm = TRUE)

gwynnda_cigs =
  trash_df |> filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |>
  pull(cigarette_butts) |> sum(na.rm = TRUE)
```

Paragraph: The combined dataset has `r nrow(trash_df)` observations from three trash wheels. Key variables include `weight_tons`, `volume_cubic_yards`, `plastic_bottles`, `cigarette_butts`, ... Professor Trash Wheel collected a total of `r prof_weight` tons of trash. Gwynnda collected `r format(gwynnda_cigs, big.mark = ",")` cigarette butts in June 2022.

# Problem 3

```{r p3_baseline, message = FALSE}
baseline_raw =
  read_csv("data/MCI_baseline.csv", skip = 1, na = c(".", "NA")) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = case_match(apoe4, 1 ~ "carrier", 0 ~ "non-carrier")
  )

# Remove participants with MCI at or before baseline
baseline_df =
  baseline_raw |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

n_recruited = nrow(baseline_raw)
n_kept = nrow(baseline_df)
n_mci = baseline_df |> filter(!is.na(age_at_onset)) |> nrow()
mean_age = mean(baseline_df$current_age)
prop_women_apoe4 =
  baseline_df |> filter(sex == "female") |>
  summarize(p = mean(apoe4 == "carrier")) |> pull(p)
```

Discussion: skipped the first description row; "." coded as missing in age at onset; recoded sex and APOE4. `r n_recruited` participants were recruited; after removing those not meeting inclusion criteria, `r n_kept` remain, and `r n_mci` developed MCI. Average baseline age is `r round(mean_age, 1)`. Proportion of women who are APOE4 carriers: `r round(prop_women_apoe4, 3)`.

```{r p3_amyloid, message = FALSE}
amyloid_df =
  read_csv("data/mci_amyloid.csv", skip = 1, na = c("NA", "Na")) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    time = str_replace(time, "time_", ""),
    time = if_else(time == "baseline", "0", time),
    time = as.numeric(time),
    amyloid_ratio = as.numeric(amyloid_ratio)
  )
```

Discussion: describe steps (skip header row, rename id, pivot longer to one row per participant-visit) and features (`r nrow(amyloid_df)` rows, `r n_distinct(amyloid_df$id)` participants).

```{r p3_join}
only_baseline = anti_join(baseline_df, amyloid_df, by = "id")
only_amyloid = anti_join(amyloid_df, baseline_df, by = "id") |> distinct(id)

combined_df = inner_join(baseline_df, amyloid_df, by = "id")

write_csv(combined_df, "data/mci_combined.csv")
```

`r nrow(only_baseline)` participants appear only in the baseline data and `r nrow(only_amyloid)` appear only in the amyloid data. The combined dataset has `r nrow(combined_df)` rows and `r n_distinct(combined_df$id)` participants.
